Long-term cloud enhancement events of global solar irradiance over Thessaloniki, Greece

### 2024-08-23 article/article.Rmd
Commit 2024-08-23 14:14

output: 
  rticles::elsevier_article:
    keep_tex:         true
    citation_package: natbib
  bookdown::pdf_document2:
    number_sections:  yes
    fig_caption:      yes
    keep_tex:         yes
    latex_engine:     xelatex
    toc:              yes
    toc_depth:        4
    fig_width:        7
    fig_height:       4.5
    documentclass:    article
    classoption:      a4paper,oneside
    fontsize:         12pt
    geometry:         "left=0.75in,right=0.75in,top=0.75in,bottom=0.75in"
    link-citations:   yes
    colorlinks:       yes
    urlcolor:         blue
    biblio-style:     apalike
  bookdown::word_document2: default
---

```{r setup, echo=F, include=FALSE}
rmarkdown::find_pandoc(dir = '/usr/lib/rstudio/resources/app/bin/quarto/bin/tools')
# renv::load("~/MANUSCRIPTS/02_enhancement/", quiet = TRUE)
## document configurations
knitr::opts_chunk$set(comment    = ""      )
# knitr::opts_chunk$set(dev        = "pdf"   )
knitr::opts_chunk$set(out.width  = "100%"   )
knitr::opts_chunk$set(fig.align  = "center")
# knitr::opts_chunk$set(fig.pos    = 'h!'    )
options(knitr.graphics.auto_pdf = FALSE)
library(pander)
library(data.table)
library(kableExtra)
library(dplyr)
## load data from trends analysis
source("~/MANUSCRIPTS/02_enhancement/GHI_enh_00_variables.R")
load("~/MANUSCRIPTS/02_enhancement/data/GHI_enh_03_process.Rda")
TSIinfo <- data.table(read.csv("~/MANUSCRIPTS/02_enhancement/figures/tbl_tsi_info.dat",
                               strip.white = TRUE,
                               sep = ";",
                               header = TRUE))
```

@@

An important aspect of the variability of solar radiation is its interaction with the
clouds. In general, clouds can attenuate a fraction of solar irradiance, but under
certain conditions, can lead to enhancement of the global horizontal irradiance (GHI)
reaching the ground. This cloud enhancement (CE) effect can locally increase the
observed GHI to levels even higher than the expected clear-sky irradiance
[@Cordero2023; @Vamvakas2020; @CastillejoCuberos2020; @Vamvakas2020 and references
therein].

@@

On multiple sites, cloud enhancements have been reported that exceed momentarily the
solar constant, resulting in clearness indices above unit. A summary of extreme cloud
enchantment (ECE) cases has been compiled by @Martins2022. Cloud enhancements can
have also some practical implications.  The intensity and duration of enhancements
can affect the efficiency and stability of photovoltaic power production
[@Lappalainen2020; @Jaervelae2020], while ECEs have the potential to compromise the
integrity of photovoltaic plants infrastructure [@DoNascimento2019].  It has also
been demonstrated that these events can interfere in the comparison of ground-based
and satellite observations of radiation [@Damiani2018].  Global warming has likely
affected cloud coverage in the last few decades. @Liu2023 reported increases in cloud
cover over the tropical and subtropical oceans and decreases over most continents,
while @Dong2023 reported decreases over North America and Europe. To our knowledge,
there is no evidence on whether this trend has affected also the number and strength
of CE events.

@@

<!-- our work -->
In this study, we evaluate the effects CE on GHI by investigating their frequency of
occurrence, intensity, and duration in a thirty-year period of GHI observations at
Thessaloniki, Greece.  We used modeled clear sky irradiance as a baseline to identify
cloud enhancements, and we determined long-term trends of the above-mentioned metrics,
their climatology and some general characteristics.  To our knowledge there are no
other studies that provide trends from such long dataset.

@@

<!-- Define CE and ECE tomove? -->
In this study, we define an event as CE when the measured GHI at ground level,
exceeds the expected value under clear-sky conditions. Similarly, we define as
extreme cloud enhancement events (ECE), the cases when GHI at ground level exceeds
the Total Solar Irradiance (TSI). Although the duration of these bursts can vary from
seconds to several minutes, here we are constrained by the temporal resolution of our
data to identify evens with duration of at minimum one-minute.

@@

This is the criterion of our CE identification.  The constant terms were determined
through the implementation of an empirical method with manual inspection of the CE
identification results on selected days of the whole dataset.  We tested seven sets
of days with different characteristics relevant to the efficiency of the
identification threshold.  These sets were random groups of about 20 -- 30 days with
the following characteristics:
(a) the strongest over irradiance CE events,
(b) the largest daily total over irradiance,
(c) absence of clouds (by implementing a clear sky identification algorithm as discussed in @Natsis2023),
(d) absence of clouds and absence of EC events,
(e) with at least $60\,\%$ of the day length without clouds and presence of EC events,
(h) randomly selected days, and
(i) manually selected days.
For the latter case and where needed, we also used images from the sky-camera to
further aid the decision of the manual inspection.

@@

Figure\nobreakspace{}\ref{fig:P-energy} shows the time series of the yearly number of
CE cases (each with duration of one minute), the yearly mean OIR and the yearly
excess irradiation for the period 1993 -- 2023, together with corresponding linear
trends. All three quantities show increasing trends, most pronounced for the
frequency of occurrence 
($`r round(yearlyCE[var == "GLB_diff.N", slope], 1)`\pm `r round(2 * yearlyCE[var == "GLB_diff.N", slope.sd], 1)`\,\text{cases}/\text{year}$)
and the excess irradiation 
($`r round(yearlyCE[var == "GLB_diff.sum", slope], 1)` \pm `r round(2 * yearlyCE[var == "GLB_diff.sum", slope.sd], 1)`\,\text{kJ}/\text{year}$),
which are also statistically significant.
In contrast the trend of the yearly mean OIR is negligible 
($`r round(yearlyCE[var == "GLB_diff.mean", slope], 1)`\pm `r round(2 * yearlyCE[var == "GLB_diff.mean", slope.sd], 1)`\,\text{W}/\text{m}^2/\text{year}$)
and of no statistical significance.
The average OIR for the entire period is
$`r round(mean(ST_E_yearly$GLB_diff.mean), 1)`\pm `r round(sd(ST_E_yearly$GLB_diff.mean),1)`\,\text{W}/\text{m}^2$.
The interannual variability of the data about the trend lines is quite large.
Furthermore, the spread tends to increase with time (at least for the quantities of
panels b and c), suggesting a significant variability in cloud patterns over the
area, possibly associated to changes in climate.

@@

The relation between the duration and the mean OIR of the groups has also been
studied (Figure\nobreakspace{}\ref{fig:group-2d}).  Evidently, evens of high excess
irradiation have small duration and vice versa. The vast majority of grouped events
are associated with small excess irradiation (e.g., $<5\,\text{kJ}/\text{m}^2$) and
small duration (e.g., $<5\,\text{min}$) while groups with strong excess irradiation
and long duration are very rare.  Similar results of this relation have been reported
by @Zhang2018 in a study using a far higher sampling rate ($100\,\text{Hz}$) than
ours.

@@

By creating a clear sky approximation representing the long- and short-term
variability of the expected clear sky GHI, we were able to identify cases of CE
events in Thessaloniki for the period 1993 -- 2023.  After analyzing the CE cases, we
found an increase of
$`r round(yearlyCE[var == "GLB_diff.N", slope], 1)`\pm `r round(2 * yearlyCE[var == "GLB_diff.N", slope.sd], 1)`\,\text{cases}/\text{year}$,
with the mean annual irradiation of the CE events increasing with a rate of
$`r round(yearlyCE[var == "GLB_diff.sum", slope], 1)`\pm `r round(2 * yearlyCE[var == "GLB_diff.sum", slope.sd], 1)`\,\text{kJ}/\text{year}$.
The most active months of CE events are May and June.  We found that continuous CE
events can last up to $140$ minutes, while the duration of $80\,\%$ of them is bellow
$5$ minutes.

>>>>

output: 
  rticles::elsevier_article:
    keep_tex:         true
    keep_md:          true
    citation_package: natbib
  bookdown::pdf_document2:
    number_sections:  yes
    fig_caption:      yes
    keep_tex:         yes
    latex_engine:     xelatex
    toc:              yes
    toc_depth:        4
    fig_width:        7
    fig_height:       4.5
    documentclass:    article
    classoption:      a4paper,oneside
    fontsize:         12pt
    geometry:         "left=0.75in,right=0.75in,top=0.75in,bottom=0.75in"
    link-citations:   yes
    colorlinks:       yes
    urlcolor:         blue
    biblio-style:     apalike
  bookdown::word_document2: default
---

```{r setup, echo=F, include=FALSE}
rmarkdown::find_pandoc(dir = '/usr/lib/rstudio/resources/app/bin/quarto/bin/tools')
# renv::load("~/MANUSCRIPTS/02_enhancement/", quiet = TRUE)
## document configurations
knitr::opts_chunk$set(comment    = ""      )
# knitr::opts_chunk$set(dev        = "pdf"   )
knitr::opts_chunk$set(out.width  = "100%"   )
knitr::opts_chunk$set(fig.align  = "center")
# knitr::opts_chunk$set(fig.pos    = 'h!'    )
options(knitr.graphics.auto_pdf = FALSE)
library(pander)
library(data.table)
library(kableExtra)
library(dplyr)
## load data from trends analysis
source("~/MANUSCRIPTS/02_enhancement/GHI_enh_00_variables.R")
load("~/MANUSCRIPTS/02_enhancement/data/GHI_enh_03_process.Rda")
TSIinfo <- data.table(read.csv("~/MANUSCRIPTS/02_enhancement/figures/tbl_tsi_info.dat",
                               strip.white = TRUE,
                               sep = ";",
                               header = TRUE))
plus_sign <- function(x) ifelse(x >= 0, paste0("+", x), x)
```

@@

An important aspect of the variability of solar radiation is its interaction with the
clouds. In general, clouds can attenuate a fraction of solar irradiance, but under
certain conditions, can lead to enhancement of the global horizontal irradiance (GHI)
reaching the ground. This cloud enhancement (CE) effect can locally increase the
observed GHI to levels even higher than the expected clear-sky irradiance
[@Cordero2023; @Vamvakas2020; @CastillejoCuberos2020;  and references
therein].

@@

On multiple sites, cloud enhancements have been reported that exceed momentarily the
solar constant, resulting in clearness indices above unit. A summary of extreme cloud
enhancement (ECE) cases has been compiled by @Martins2022. Cloud enhancements can
have also some practical implications.  The intensity and duration of enhancements
can affect the efficiency and stability of photovoltaic power production
[@Lappalainen2020; @Jaervelae2020], while ECEs have the potential to compromise the
integrity of photovoltaic plants infrastructure [@DoNascimento2019].  It has also
been demonstrated that these events can interfere in the comparison of ground-based
and satellite observations of radiation [@Damiani2018].  Global warming has likely
affected cloud coverage in the last few decades. @Liu2023 reported increases in cloud
cover over the tropical and subtropical oceans and decreases over most continents,
while @Dong2023 reported decreases over North America and Europe. To our knowledge,
there is no evidence on whether this trend has affected also the number and strength
of CE events.

@@

<!-- our work -->
In this study, we evaluate the effects of CE on GHI by investigating their frequency
of occurrence, intensity, and duration in a thirty-year period of GHI observations at
Thessaloniki, Greece.  We used modeled clear sky irradiance as a baseline to identify
cloud enhancements, and we determined long-term trends of the above-mentioned
metrics, their climatology and some general characteristics.  To our knowledge there
are no other studies that provide trends from such long dataset.

@@

<!-- Define CE and ECE tomove? -->
In this study, we define an event as CE when the measured GHI at ground level,
exceeds the expected value under clear-sky conditions. Similarly, we define as
extreme cloud enhancement events (ECE), the cases when GHI at ground level exceeds
the Total Solar Irradiance (TSI). Although the duration of these bursts can vary from
seconds to several minutes, here we are constrained by the temporal resolution of our
data to identify events with duration of at minimum one-minute.

@@

This is the criterion of our CE identification.  The constant terms were determined
through the implementation of an empirical method with manual inspection of the CE
identification results on selected days of the whole dataset.  We tested seven sets
of days with different characteristics relevant to the efficiency of the
identification threshold.  These sets were random groups of about 20 -- 30 days with
the following characteristics:
(a) the strongest over irradiance CE events,
(b) the largest daily total over irradiance,
(c) absence of clouds (by implementing a clear sky identification algorithm as discussed in @Natsis2023),
(d) absence of clouds and absence of CE events,
(e) with at least $60\,\%$ of the day length without clouds and presence of CE events,
(h) randomly selected days, and
(i) manually selected days.
For the latter case and where needed, we also used images from the sky-camera to
further aid the decision of the manual inspection.

@@

Figure\nobreakspace{}\ref{fig:P-energy} shows the time series of the yearly number of
CE cases (each with duration of one minute), the yearly mean OIR and the yearly
excess irradiation for the period 1993 -- 2023, together with corresponding linear
trends. All three quantities show increasing trends, most pronounced for the
frequency of occurrence 
($`r plus_sign(round(yearlyCE[var == "GLB_diff.N", slope], 1))`\pm `r round(2 * yearlyCE[var == "GLB_diff.N", slope.sd], 1)`\,\text{cases}/\text{year}$)
and the excess irradiation 
($`r plus_sign(round(yearlyCE[var == "GLB_diff.sum", slope], 1))` \pm `r round(2 * yearlyCE[var == "GLB_diff.sum", slope.sd], 1)`\,\text{kJ}/\text{year}$),
which are also statistically significant.
In contrast the trend of the yearly mean OIR is negligible 
($`r plus_sign(round(yearlyCE[var == "GLB_diff.mean", slope], 1))`\pm `r round(2 * yearlyCE[var == "GLB_diff.mean", slope.sd], 1)`\,\text{W}/\text{m}^2/\text{year}$)
and of no statistical significance.
The average OIR for the entire period is
$`r plus_sign(round(mean(ST_E_yearly$GLB_diff.mean), 1))`\pm `r round(sd(ST_E_yearly$GLB_diff.mean),1)`\,\text{W}/\text{m}^2$.
The interannual variability of the data about the trend lines is quite large.
Furthermore, the spread tends to increase with time (at least for the quantities of
panels b and c), suggesting a significant variability in cloud patterns over the
area, possibly associated to changes in climate.

@@

The relation between the duration and the mean OIR of the groups has also been
studied (Figure\nobreakspace{}\ref{fig:group-2d}).  Evidently, events of high excess
irradiation have small duration and vice versa. The vast majority of grouped events
are associated with small excess irradiation (e.g., $<5\,\text{kJ}/\text{m}^2$) and
small duration (e.g., $<5\,\text{min}$) while groups with strong excess irradiation
and long duration are very rare.  Similar results of this relation have been reported
by @Zhang2018 in a study using a far higher sampling rate ($100\,\text{Hz}$) than
ours.

@@

By creating a clear sky approximation representing the long- and short-term
variability of the expected clear sky GHI, we were able to identify cases of CE
events in Thessaloniki for the period 1993 -- 2023.  After analyzing the CE cases, we
found an increase of
$`r plus_sign(round(yearlyCE[var == "GLB_diff.N", slope], 1))`\pm `r round(2 * yearlyCE[var == "GLB_diff.N", slope.sd], 1)`\,\text{cases}/\text{year}$,
with the mean annual irradiation of the CE events increasing with a rate of
$`r plus_sign(round(yearlyCE[var == "GLB_diff.sum", slope], 1))`\pm `r round(2 * yearlyCE[var == "GLB_diff.sum", slope.sd], 1)`\,\text{kJ}/\text{year}$.
The most active months of CE events are May and June.  We found that continuous CE
events can last up to $140$ minutes, while the duration of $80\,\%$ of them is bellow
$5$ minutes.

### 2024-08-23 article/article.tex
Commit 2024-08-23 14:14

An important aspect of the variability of solar radiation is its
interaction with the clouds. In general, clouds can attenuate a fraction
of solar irradiance, but under certain conditions, can lead to
enhancement of the global horizontal irradiance (GHI) reaching the
ground. This cloud enhancement (CE) effect can locally increase the
observed GHI to levels even higher than the expected clear-sky
irradiance \citep[ and references
therein]{Cordero2023, Vamvakas2020, CastillejoCuberos2020, Vamvakas2020}.

@@

On multiple sites, cloud enhancements have been reported that exceed
momentarily the solar constant, resulting in clearness indices above
unit. A summary of extreme cloud enchantment (ECE) cases has been
compiled by \citet{Martins2022}. Cloud enhancements can have also some
practical implications. The intensity and duration of enhancements can
affect the efficiency and stability of photovoltaic power production
\citep{Lappalainen2020, Jaervelae2020}, while ECEs have the potential to
compromise the integrity of photovoltaic plants infrastructure
\citep{DoNascimento2019}. It has also been demonstrated that these
events can interfere in the comparison of ground-based and satellite
observations of radiation \citep{Damiani2018}. Global warming has likely
affected cloud coverage in the last few decades. \citet{Liu2023}
reported increases in cloud cover over the tropical and subtropical
oceans and decreases over most continents, while \citet{Dong2023}
reported decreases over North America and Europe. To our knowledge,
there is no evidence on whether this trend has affected also the number
and strength of CE events.

@@

In this study, we evaluate the effects CE on GHI by investigating their
frequency of occurrence, intensity, and duration in a thirty-year period
of GHI observations at Thessaloniki, Greece. We used modeled clear sky
irradiance as a baseline to identify cloud enhancements, and we
determined long-term trends of the above-mentioned metrics, their
climatology and some general characteristics. To our knowledge there are
no other studies that provide trends from such long dataset.

@@

In this study, we define an event as CE when the measured GHI at ground
level, exceeds the expected value under clear-sky conditions. Similarly,
we define as extreme cloud enhancement events (ECE), the cases when GHI
at ground level exceeds the Total Solar Irradiance (TSI). Although the
duration of these bursts can vary from seconds to several minutes, here
we are constrained by the temporal resolution of our data to identify
evens with duration of at minimum one-minute.

@@

This is the criterion of our CE identification. The constant terms were
determined through the implementation of an empirical method with manual
inspection of the CE identification results on selected days of the
whole dataset. We tested seven sets of days with different
characteristics relevant to the efficiency of the identification
threshold. These sets were random groups of about 20 -- 30 days with the
following characteristics: (a) the strongest over irradiance CE events,
(b) the largest daily total over irradiance, (c) absence of clouds (by
implementing a clear sky identification algorithm as discussed in
\citet{Natsis2023}), (d) absence of clouds and absence of EC events, (e)
with at least \(60\,\%\) of the day length without clouds and presence
of EC events, (h) randomly selected days, and (i) manually selected
days. For the latter case and where needed, we also used images from the
sky-camera to further aid the decision of the manual inspection.

@@

Figure\nobreakspace{}\ref{fig:P-energy} shows the time series of the
yearly number of CE cases (each with duration of one minute), the yearly
mean OIR and the yearly excess irradiation for the period 1993 -- 2023,
together with corresponding linear trends. All three quantities show
increasing trends, most pronounced for the frequency of occurrence
(\(45.6\pm 21.9\,\text{cases}/\text{year}\)) and the excess irradiation
(\(116.9 \pm 67.8\,\text{kJ}/\text{year}\)), which are also
statistically significant. In contrast the trend of the yearly mean OIR
is negligible (\(0.1\pm 0.1\,\text{W}/\text{m}^2/\text{year}\)) and of
no statistical significance. The average OIR for the entire period is
\(39.9\pm 2.7\,\text{W}/\text{m}^2\). The interannual variability of the
data about the trend lines is quite large. Furthermore, the spread tends
to increase with time (at least for the quantities of panels b and c),
suggesting a significant variability in cloud patterns over the area,
possibly associated to changes in climate.

@@

\caption{Relative frequency distribution of CE groups of consequent CE events according to their duration for Thessaloniki in the period 1993 -- 2023.}\label{fig:ceg-duration-distribution}
\end{figure}

The relation between the duration and the mean OIR of the groups has
also been studied (Figure\nobreakspace{}\ref{fig:group-2d}). Evidently,
evens of high excess irradiation have small duration and vice versa. The
vast majority of grouped events are associated with small excess
irradiation (e.g., \(<5\,\text{kJ}/\text{m}^2\)) and small duration
(e.g., \(<5\,\text{min}\)) while groups with strong excess irradiation
and long duration are very rare. Similar results of this relation have
been reported by \citet{Zhang2018} in a study using a far higher
sampling rate (\(100\,\text{Hz}\)) than ours.

@@

By creating a clear sky approximation representing the long- and
short-term variability of the expected clear sky GHI, we were able to
identify cases of CE events in Thessaloniki for the period 1993 -- 2023.
After analyzing the CE cases, we found an increase of
\(45.6\pm 21.9\,\text{cases}/\text{year}\), with the mean annual
irradiation of the CE events increasing with a rate of
\(116.9\pm 67.8\,\text{kJ}/\text{year}\). The most active months of CE
events are May and June. We found that continuous CE events can last up
to \(140\) minutes, while the duration of \(80\,\%\) of them is bellow
\(5\) minutes.

>>>>

An important aspect of the variability of solar radiation is its
interaction with the clouds. In general, clouds can attenuate a fraction
of solar irradiance, but under certain conditions, can lead to
enhancement of the global horizontal irradiance (GHI) reaching the
ground. This cloud enhancement (CE) effect can locally increase the
observed GHI to levels even higher than the expected clear-sky
irradiance {[}\citet{Cordero2023}; \citet{Vamvakas2020};
\citet{CastillejoCuberos2020}; and references therein{]}.

@@

On multiple sites, cloud enhancements have been reported that exceed
momentarily the solar constant, resulting in clearness indices above
unit. A summary of extreme cloud enhancement (ECE) cases has been
compiled by \citet{Martins2022}. Cloud enhancements can have also some
practical implications. The intensity and duration of enhancements can
affect the efficiency and stability of photovoltaic power production
\citep{Lappalainen2020, Jaervelae2020}, while ECEs have the potential to
compromise the integrity of photovoltaic plants infrastructure
\citep{DoNascimento2019}. It has also been demonstrated that these
events can interfere in the comparison of ground-based and satellite
observations of radiation \citep{Damiani2018}. Global warming has likely
affected cloud coverage in the last few decades. \citet{Liu2023}
reported increases in cloud cover over the tropical and subtropical
oceans and decreases over most continents, while \citet{Dong2023}
reported decreases over North America and Europe. To our knowledge,
there is no evidence on whether this trend has affected also the number
and strength of CE events.

@@

In this study, we evaluate the effects of CE on GHI by investigating
their frequency of occurrence, intensity, and duration in a thirty-year
period of GHI observations at Thessaloniki, Greece. We used modeled
clear sky irradiance as a baseline to identify cloud enhancements, and
we determined long-term trends of the above-mentioned metrics, their
climatology and some general characteristics. To our knowledge there are
no other studies that provide trends from such long dataset.

@@

In this study, we define an event as CE when the measured GHI at ground
level, exceeds the expected value under clear-sky conditions. Similarly,
we define as extreme cloud enhancement events (ECE), the cases when GHI
at ground level exceeds the Total Solar Irradiance (TSI). Although the
duration of these bursts can vary from seconds to several minutes, here
we are constrained by the temporal resolution of our data to identify
events with duration of at minimum one-minute.

@@

This is the criterion of our CE identification. The constant terms were
determined through the implementation of an empirical method with manual
inspection of the CE identification results on selected days of the
whole dataset. We tested seven sets of days with different
characteristics relevant to the efficiency of the identification
threshold. These sets were random groups of about 20 -- 30 days with the
following characteristics: (a) the strongest over irradiance CE events,
(b) the largest daily total over irradiance, (c) absence of clouds (by
implementing a clear sky identification algorithm as discussed in
\citet{Natsis2023}), (d) absence of clouds and absence of CE events, (e)
with at least \(60\,\%\) of the day length without clouds and presence
of CE events, (h) randomly selected days, and (i) manually selected
days. For the latter case and where needed, we also used images from the
sky-camera to further aid the decision of the manual inspection.

@@

Figure\nobreakspace{}\ref{fig:P-energy} shows the time series of the
yearly number of CE cases (each with duration of one minute), the yearly
mean OIR and the yearly excess irradiation for the period 1993 -- 2023,
together with corresponding linear trends. All three quantities show
increasing trends, most pronounced for the frequency of occurrence
(\(+45.6\pm 21.9\,\text{cases}/\text{year}\)) and the excess irradiation
(\(+116.9 \pm 67.8\,\text{kJ}/\text{year}\)), which are also
statistically significant. In contrast the trend of the yearly mean OIR
is negligible (\(+0.1\pm 0.1\,\text{W}/\text{m}^2/\text{year}\)) and of
no statistical significance. The average OIR for the entire period is
\(+39.9\pm 2.7\,\text{W}/\text{m}^2\). The interannual variability of
the data about the trend lines is quite large. Furthermore, the spread
tends to increase with time (at least for the quantities of panels b and
c), suggesting a significant variability in cloud patterns over the
area, possibly associated to changes in climate.

@@

\caption{Relative frequency distribution of CE groups of consequent CE events according to their duration for Thessaloniki in the period 1993 -- 2023. The histogram was split in two plots with different y-axis scales for better readability.}\label{fig:ceg-duration-distribution}
\end{figure}

The relation between the duration and the mean OIR of the groups has
also been studied (Figure\nobreakspace{}\ref{fig:group-2d}). Evidently,
events of high excess irradiation have small duration and vice versa.
The vast majority of grouped events are associated with small excess
irradiation (e.g., \(<5\,\text{kJ}/\text{m}^2\)) and small duration
(e.g., \(<5\,\text{min}\)) while groups with strong excess irradiation
and long duration are very rare. Similar results of this relation have
been reported by \citet{Zhang2018} in a study using a far higher
sampling rate (\(100\,\text{Hz}\)) than ours.

@@

By creating a clear sky approximation representing the long- and
short-term variability of the expected clear sky GHI, we were able to
identify cases of CE events in Thessaloniki for the period 1993 -- 2023.
After analyzing the CE cases, we found an increase of
\(+45.6\pm 21.9\,\text{cases}/\text{year}\), with the mean annual
irradiation of the CE events increasing with a rate of
\(+116.9\pm 67.8\,\text{kJ}/\text{year}\). The most active months of CE
events are May and June. We found that continuous CE events can last up
to \(140\) minutes, while the duration of \(80\,\%\) of them is bellow
\(5\) minutes.
